A properly labelled proof system for intuitionistic modal logics

### 2019-09-16 caleidoscope/caleidoscope.tex
working on soundness section

We write $\M, w \Vdash a$ if $a \in V(w)$, and by definition, we always have $\M, w \Vdash \top$ and never that $\M, w \Vdash \bot$. 
%
Then the relation is extended to all formulas by induction, following the rules for both intuitionistic and modal Kripke models:
$$
\begin{array}{rcl}
\label{eq:kripke}
%\parbox{30em}{%

@@

In this section we present our properly labelled sequent proof system
for intuitionistic modal logics. The starting point is the notion of a
\emph{labelled formula} which is a pair $\labels xA$ of a label $\lb
x$ and a formula $\fm A$. The intended meaning is that $\fm A$ holds
at world $\lb x$. A \emph{relation atom} is either an expression
$\accs xy$ or $\futs xy$ where $\lb x$ and $\lb y$ are labels and the
intended meaning is that the corresponing relation holds between the
worlds $\lb x$ and $\lb y$. This is enough to define the notion of
\emph{(labelled) sequent}, which is a triple $\lseq\B\Left\Right$,
where $\B$ is a set of relation atoms, and $\Left$ and $\Right$ are
multi-sets of labelled formulas, all written as list, separated by commas.

\begin{figure}[!t]
  \begin{center}
	\fbox{
%	\begin{minipage}{.95\textwidth}
	\begin{tabular}{c@{\quad}c}
		$\vlinf{\rn{id}}{}{\B, x \le y, \Left, \labels{x}{a} \SEQ \Right, \labels{y}{a} }{}$
		&
		$\vlinf{\llabrn\bot}{}{\B, \Left, \labels{x}{\BOT} \SEQ \Right}{}$
		%\quad
		%$\vlinf{\rlabrn\top}{}{\B, \Left \SEQ \Right, \labels{x}{\TOP}}{}$
		\\\\
		$\vlinf{\llabrn\AND}{}{\B,\Left, \labels{x}{A \AND B} \SEQ \Right}{\B, \Left, \labels{x}{A}, \labels{x}{B} \SEQ \Right}$
		&
		$\vliinf{\rlabrn\AND}{}{\B,\Left \SEQ \Right, \labels{x}{A \AND B}}{\B, \Left \SEQ \Right, \labels{x}{A}}{\B, \Left \SEQ \Right, \labels{x}{B}}$
		\\\\
		$\vliinf{\llabrn\OR}{}{\B, \Left, \labels{x}{A \OR B} \SEQ \Right}{\B, \Left, \labels{x}{A} \SEQ \Right}{\B, \Left, \labels{x}{B} \SEQ \Right}$
		&
		$\vlinf{\rlabrn\OR}{}{\B, \Left \SEQ \Right, \labels{x}{A \OR B}}{\B, \Left \SEQ \Right, \labels{x}{A}, \labels{x}{B}}$
		\\\\
		\multicolumn{2}{c}{
		$\vlinf{\rlabrn\IMP}{\text{\scriptsize $y$ fresh}}{\B, \Left \SEQ \Right, \labels{x}{A \IMP B}}{\B, x \le y, \Left, \labels{y}{A} \SEQ \Right, \labels{y}{B}}$
		}
		\\\\
		\multicolumn{2}{c}{
 		$\vliinf{\llabrn\IMP}{}{\B, x \le y, \Left, \labels{x}{A \IMP B} \SEQ \Right}{\B, x \le y, \labels{x}{A \IMP B}, \Left \SEQ \Right, \labels{y}{A}}{\B, x \le y, \Left, \labels{y}{B} \SEQ \Right}$ 
% 		&$\vliinf{\rlabrn\IMP}{\text{\scriptsize $x \le y \in \B$}}{\B, \Left, \labels{x}{A \IMP B} \SEQ \Right}{\B, \Left \SEQ \Right, \labels{y}{A}}{\B, \Left, \labels{y}{B} \SEQ \Right}$
 		}
		\\\\
		$\vlinf{\llabrn\BOX}{}{\B, x \le y, yRz, \Left, \labels{x}{\BOX A} \SEQ \Right}{\B, x \le y, yRz, \Left, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right}$
		&
		$\vlinf{\rlabrn\BOX}{\text{\scriptsize $y, z$ fresh}}{\B, \Left \SEQ \Right, \labels{x}{\BOX A}}{\B, x \le y, y \rel z, \Left \SEQ \Right, \labels{z}{A}}$
		\\\\
		$\vlinf{\llabrn\DIA}{\text{\scriptsize $y$ fresh}}{\B, \Left, \labels{x}{\DIA A} \SEQ \Right}{\B, x \rel y, \Left, \labels{y}{A} \SEQ \Right}$
		&
		$\vlinf{\rlabrn\DIA}{}{\B, x \rel y, \Left \SEQ \Right, \labels{x}{\DIA A}}{\B, x \rel y, \Left \SEQ \Right, \labels{x}{\DIA A}, \labels{y}{A}}$
		\\
		\multicolumn{2}{c}{
		$\mbox{\hbox to .9\linewidth{\dotfill}}$
		}
		\\
		$\vlinf{\rn{refl}}{}{\B, \Left \SEQ \Right}{\B, x\le x, \Left \SEQ \Right}$
		&
		$\vlinf{\rn{trans}}{}{\B, x \le y, y \le z, \Left \SEQ \Right}{\B, x \le y, y \le z, x \le z, \Left \SEQ \Right}$
		\\\\
		\multicolumn{2}{c}{
		$\vlinf{\rn{F_1}}{\text{\scriptsize $u$ fresh}}{\B, xRy, y \le z, \Left \SEQ \Right}{\B, xRy, y \le z, x \le u, uRz, \Left \SEQ \Right}$
		}
		\\\\
		\multicolumn{2}{c}{
		$\vlinf{\rn{F_2}}{\text{\scriptsize $u$ fresh}}{\B, xRy,x \le z, \Left \SEQ \Right}{\B, xRy, x \le z, y \le u, zRu, \Left \SEQ \Right }$		
		}
	\end{tabular}		
%\end{minipage}
}		
  \end{center}
  \caption{System $\labIKp$}
	\label{fig:labIKp}
\end{figure}

@@

\todo{}

@@

@@

\begin{theorem}
	For any formula $A$. If $A$ is a theorem of $\IK$ then $A$ is provable in $\labIKp +\labrn{cut}$.
\end{theorem}

@@

	\begin{center}
		$\vlderivation{\vlin {\llabrn{\OR}}
			{}
			{\star}
			{\vlin {\rlabrn{\OR}}
				{}
				{x \le y, yRz, \labels{z}{B} \Rightarrow \labels{y}{\DIA A \OR \DIA B}}
				{\vlin {\rlabrn{\DIA}}
					{}
					{x \le y, yRz, \labels{z}{B} \Rightarrow \labels{y}{\DIA A}, \labels{y}{\DIA B}}
					{\vlin {\rn{refl}}
						{}
						{x \le y, yRz, \labels{z}{B} \Rightarrow \labels{y}{\DIA A}, \labels{y}{\DIA B}, \labels{z}{B}}
						{\vlin {\rn{id}}
							{}
							{x \le y, z \le z, yRz, \labels{z}{B} \Rightarrow \labels{y}{\DIA A}, \labels{y}{\DIA B}, \labels{z}{B}}
							{\vlhy {}}}}}}}$
	\end{center}
\end{proof}

@@

\begin{center}
	$\vlderivation {
		\vlin{\rlabrn{\IMP}}
		{}
		{\Rightarrow \labels{x}{A \AND B \IMP B}}
		{\vlin {\llabrn{\AND}}
			{}
			{x \le y, \labels{y}{A \AND B} \Rightarrow \labels{y}{B}}
			{\vlin {\rn{refl}}
				{}
				{x \le y, \labels{y}{A}, \labels{y}{B} \Rightarrow \labels{y}{B}}
				{\vlin {\rn{id}}
					{}
					{x \le y, y \le y, \labels{y}{A}, \labels{y}{B} \Rightarrow \labels{y}{B}}
					{\vlhy {}}}}}
	}$
\end{center}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{lemma}\label{lem:inv}\hbox{}\quad
	\begin{enumerate}
		\item
		If there exists a proof $\vlderivation{\vlhtr{\DD}{\B, \Left, \labels{x}{A \AND B} \SEQ \Right}}$ then there exists a proof $\vlderivation{\vlhtr{\DD^{\invr{\llabrn\AND}}}{\B, \Left, \labels{x}{A}, \labels{x}{B} \SEQ \Right}}$ of the same (or smaller) height, for any label $x$ that does not appear in $\DD$.

@@

>>>>

We write $\M, \lb w \Vdash a$ if $a \in V(w)$, and by definition, we always have $\M, w \Vdash \top$ and never that $\M, w \Vdash \bot$. 
%
Then the relation is extended to all formulas by induction, following the rules for both intuitionistic and modal Kripke models:
$$
\begin{array}{rcl}
\label{eq:kripke}
%\parbox{30em}{%

@@

In this section we present our properly labelled sequent proof system
for intuitionistic modal logics. The starting point is the notion of a
\emph{labelled formula} which is a pair $\labels xA$ of a label $\lb
x$ and a formula $\fm A$. The intended meaning is that $\fm A$ holds
at world $\lb x$. A \emph{relation atom} is either an expression
$\accs xy$ or $\futs xy$ where $\lb x$ and $\lb y$ are labels and the
intended meaning is that the corresponing relation holds between the
worlds $\lb x$ and $\lb y$. This is enough to define the notion of
\emph{(labelled) sequent}, which is a triple $\lseq\B\Left\Right$,
where $\B$ is a set of relational and pre-order atoms, and $\Left$ and $\Right$ are
multi-sets of labelled formulas, all written as list, separated by commas.

\begin{figure}[!t]
  \begin{center}
	\fbox{
%	\begin{minipage}{.95\textwidth}
	\begin{tabular}{c@{\quad}c}
		$\vlinf{\rn{id}}{}{\B, \futs xy, \Left, \labels{x}{a} \SEQ \Right, \labels{y}{a} }{}$
		&
		$\vlinf{\llabrn\bot}{}{\B, \Left, \labels{x}{\BOT} \SEQ \Right}{}$
		%\quad
		%$\vlinf{\rlabrn\top}{}{\B, \Left \SEQ \Right, \labels{x}{\TOP}}{}$
		\\\\
		$\vlinf{\llabrn\AND}{}{\B,\Left, \labels{x}{A \AND B} \SEQ \Right}{\B, \Left, \labels{x}{A}, \labels{x}{B} \SEQ \Right}$
		&
		$\vliinf{\rlabrn\AND}{}{\B,\Left \SEQ \Right, \labels{x}{A \AND B}}{\B, \Left \SEQ \Right, \labels{x}{A}}{\B, \Left \SEQ \Right, \labels{x}{B}}$
		\\\\
		$\vliinf{\llabrn\OR}{}{\B, \Left, \labels{x}{A \OR B} \SEQ \Right}{\B, \Left, \labels{x}{A} \SEQ \Right}{\B, \Left, \labels{x}{B} \SEQ \Right}$
		&
		$\vlinf{\rlabrn\OR}{}{\B, \Left \SEQ \Right, \labels{x}{A \OR B}}{\B, \Left \SEQ \Right, \labels{x}{A}, \labels{x}{B}}$
		\\\\
		\multicolumn{2}{c}{
		$\vlinf{\rlabrn\IMP}{\text{\scriptsize $y$ fresh}}{\B, \Left \SEQ \Right, \labels{x}{A \IMP B}}{\B, \futs xy, \Left, \labels{y}{A} \SEQ \Right, \labels{y}{B}}$
		}
		\\\\
		\multicolumn{2}{c}{
 		$\vliinf{\llabrn\IMP}{}{\B, \futs xy, \Left, \labels{x}{A \IMP B} \SEQ \Right}{\B, \futs xy, \labels{x}{A \IMP B}, \Left \SEQ \Right, \labels{y}{A}}{\B, \futs xy, \Left, \labels{y}{B} \SEQ \Right}$ 
% 		&$\vliinf{\rlabrn\IMP}{\text{\scriptsize $x \le y \in \B$}}{\B, \Left, \labels{x}{A \IMP B} \SEQ \Right}{\B, \Left \SEQ \Right, \labels{y}{A}}{\B, \Left, \labels{y}{B} \SEQ \Right}$
 		}
		\\\\
		$\vlinf{\llabrn\BOX}{}{\B, \futs xy, \accs yz, \Left, \labels{x}{\BOX A} \SEQ \Right}{\B, x \le y, \accs yz, \Left, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right}$
		&
		$\vlinf{\rlabrn\BOX}{\text{\scriptsize $y, z$ fresh}}{\B, \Left \SEQ \Right, \labels{x}{\BOX A}}{\B, \futs xy, \accs yz, \Left \SEQ \Right, \labels{z}{A}}$
		\\\\
		$\vlinf{\llabrn\DIA}{\text{\scriptsize $y$ fresh}}{\B, \Left, \labels{x}{\DIA A} \SEQ \Right}{\B, \accs xy, \Left, \labels{y}{A} \SEQ \Right}$
		&
		$\vlinf{\rlabrn\DIA}{}{\B, \accs xy, \Left \SEQ \Right, \labels{x}{\DIA A}}{\B, \accs xy, \Left \SEQ \Right, \labels{x}{\DIA A}, \labels{y}{A}}$
		\\
		\multicolumn{2}{c}{
		$\mbox{\hbox to .9\linewidth{\dotfill}}$
		}
		\\
		$\vlinf{\rn{refl}}{}{\B, \Left \SEQ \Right}{\B, \futs xx, \Left \SEQ \Right}$
		&
		$\vlinf{\rn{trans}}{}{\B, \futs xy, \futs yz, \Left \SEQ \Right}{\B, \futs xy, \futs yz, \futs xz, \Left \SEQ \Right}$
		\\\\
		\multicolumn{2}{c}{
		$\vlinf{\rn{F_1}}{\text{\scriptsize $u$ fresh}}{\B, \accs xy, \futs yz, \Left \SEQ \Right}{\B, \accs xy, \futs yz, \futs xu, \accs uz, \Left \SEQ \Right}$
		}
		\\\\
		\multicolumn{2}{c}{
		$\vlinf{\rn{F_2}}{\text{\scriptsize $u$ fresh}}{\B, \accs xy, \futs xz, \Left \SEQ \Right}{\B, \accs xy, \futs xz, \futs yu, \accs zu, \Left \SEQ \Right }$		
		}
	\end{tabular}		
%\end{minipage}
}		
  \end{center}
  \caption{System $\labIKp$}
	\label{fig:labIKp}
\end{figure}

@@

\todo{move lemma \ref{lem:inv}}

@@

In order to prove the implication \ref{iii}$\implies$\ref{iv} from Theorem~\ref{thm:cutfree-compl} we need to show that each sequent rule of our system $\labIKp$ is sound. We want to prove that for all model $\M$ that satisfies the premise, then $\M$ satifies the conclusion.

In order to obtain the proof, let us introduce the next definitions:

\begin{definition}
	Let $\M = \langle W, \rel, \le, V \rangle$ be a model. We say that $\f$ is an \emph{assignment function} such that $\f : Labels \rightarrow W$.
\end{definition}

\begin{definition}
	Let $\M = \langle W, \rel, \le, V \rangle$ be a model and $\B, \Gamma \Rightarrow \Delta$ a sequent. We say that $\M \Vdash \B, \Gamma \Rightarrow \Delta$ if there exists $\f$ such that:
	\begin{enumerate}
		\item For all $\labels{x}{A} \in \Gamma$, we have $\M, f(\lb x) \Vdash \fm A$.
		\item For all $\lb x, \lb y \in \B$, we have $f(\lb x) \rel f(\lb y)$.
		\item For all $\futs xy \in \B$, we have $f(\lb x) \le f(\lb y)$.
	\end{enumerate}
	then for all $\labels{z}{B} \in \Delta$, we have that $\M, f(\lb z) \Vdash \fm B$. We say that $\M \not \Vdash \B, \Gamma \Rightarrow \Delta$ if it is not the case that $\M \Vdash \B, \Gamma \Rightarrow \Delta$.
\end{definition}

Following the definitions, we display the proof for some rules since it is similar for each rule. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\todo{an introduction? do not repeat the theorem?}
\begin{theorem}
	For any formula $A$. If $A$ is a theorem of $\IK$ then $A$ is provable in $\labIKp +\labrn{cut}$.
\end{theorem}

@@

	\begin{center}
		$\vlderivation{\vlin {\llabrn{\OR}}
			{}
			{\star}
			{\vlin {\rlabrn{\OR}}
				{}
				{x \le y, yRz, \labels{z}{B} \Rightarrow \labels{y}{\DIA A \OR \DIA B}}
				{\vlin {\rlabrn{\DIA}}
					{}
					{x \le y, yRz, \labels{z}{B} \Rightarrow \labels{y}{\DIA A}, \labels{y}{\DIA B}}
					{\vlin {\rn{refl}}
						{}
						{x \le y, yRz, \labels{z}{B} \Rightarrow \labels{y}{\DIA A}, \labels{y}{\DIA B}, \labels{z}{B}}
						{\vlin {\rn{id}}
							{}
							{x \le y, z \le z, yRz, \labels{z}{B} \Rightarrow \labels{y}{\DIA A}, \labels{y}{\DIA B}, \labels{z}{B}}
							{\vlhy {}}}}}}}$
	\end{center}

@@

\begin{center}
	$\vlderivation {
		\vlin{\rlabrn{\IMP}}
		{}
		{\Rightarrow \labels{x}{A \AND B \IMP B}}
		{\vlin {\llabrn{\AND}}
			{}
			{x \le y, \labels{y}{A \AND B} \Rightarrow \labels{y}{B}}
			{\vlin {\rn{refl}}
				{}
				{x \le y, \labels{y}{A}, \labels{y}{B} \Rightarrow \labels{y}{B}}
				{\vlin {\rn{id}}
					{}
					{x \le y, y \le y, \labels{y}{A}, \labels{y}{B} \Rightarrow \labels{y}{B}}
					{\vlhy {}}}}}
	}$
\end{center}

\end{proof}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{lemma}\label{lem:inv}\hbox{}\quad
	\begin{enumerate}
		%\item
		%If there exists a proof $\vlderivation{\vlhtr{\DD}{\B, \Left, \labels{x}{A \AND B} \SEQ \Right}}$ then there exists a proof $\vlderivation{\vlhtr{\DD^{\invr{\llabrn\AND}}}{\B, \Left, \labels{x}{A}, \labels{x}{B} \SEQ \Right}}$ of the same (or smaller) height, for any label $x$ that does not appear in $\DD$.

@@

\todo{add missing cases}
